Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-16 manuscript/manuscript.Rmd
more ms updates; moves discussion opener to intro

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual-level data we fit models for survival, growth, and recruitment for each species. At the quadrat-level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions. Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.

\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}

>>>>

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual-level data we fit models for survival, growth, and recruitment for each species. At the quadrat-level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariates that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and April through June temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions. Climate $\times$ size interactions are for climate main effects only, that is we do not include interactions among size and interacting climate effects.

### 2015-07-16 manuscript/manuscript.tex
more ms updates; moves discussion opener to intro

To date, most empirical population modelers have relied on
individual-level data, with few attempts to exploit population-level
data. An important exception was an effort by Taylor and Hastings (2004)
to model the population growth rate of an invasive species to identify
the best strategies for invasion control. They used a
``density-structured'' model where the state variable is a discrete
density state rather than a continuous density measure. Building on this
work, Freckleton et al. (2011) showed that density-structured models
compare well to continuous models in theory, and Queenborough et al.
(2011) demonstrated the application of such methods in a study on arable
weeds. In particular, Queenborough et al. (2011) provide empirical
evidence that density-structured models are capable of reproducing
population dynamics, even if some precision is lost when compared to
fully continuous models. Thus, population models based on coarse,
population-level data show promise for producing ecological forecasts at
landscape and regional scales (Queenborough et al. 2011). However, none
of these models included environmental covariates.

@@

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual-level data we fit models for survival, growth, and
recruitment for each species. At the quadrat-level we fit a single
regression model for population growth. We describe the statistical
models separately since fitting the models required different
approaches. All models contain five climate covariate that we chose
\emph{a priori}: ``water year'' precipitation at \emph{t}-1 (lagppt);
fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year. We also include interactions among same-year
climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
size interactions. Climate $\times$ size interactions are for climate
main effects only, that is we do not include interactions among size and
interacting climate effects.

\todo{let's talk about timing of "spring". Miles City gets a lot of moisture in May and June...}

We fit all models using a hierarchical Bayesian approach. The models are
fully descibed in Appendix A, so here we focus on the main process and
the model likelihood. For the likelihood models, $\textbf{y}^X$ is
always the relevant vector of observations for vital rate \emph{X}
($X = S, G, R, or P for survival, growth, recruitment, or population growth). For example, $\textbf{y}\^{}S\$
is a vector of 0s and 1s indicating whether a genet survives from
\emph{t} to \emph{t+1}, or not.

@@

which means we are calculating growth (\emph{G}) for individuals that
survive (\emph{S}) from time \emph{t} to \emph{t+1} and adding in newly
recruited (\emph{R}) individuals of an average sized one-year-old genet
for the focal species. Our stastical model for recruitment (\emph{R},
described above) returns the number of new recruit produced per quadrat.
Following previous work (Adler et al. 2012, Chu and Adler 2015), we
assume that fecundity increases linearly with size
($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
to incorporate the recruitment function in the spatially-implicit IPM.
{[}You probably should say something about spatial structure and cite
Adler et al. 2010 EcoLetts for the ``no overlap rule''{]}

@@

Our quad-based model (QBM) perfectly mirrors {[}I think you need to say
a little more about how you simulate it{]} its statistical description
(Eqs. 8-9). We use the same approach for drawing parameter values as
described for the IPM.

@@

@@

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Species showed similar trends (Fig. 3e-h). Growth was the
most sensitive vital rate for all species, showing a negative response
to increased precipitation, and stronger positive response to increased
temperature, and a mostly positive response when both climate factors
are increased (Fig. 3e-h). \emph{B. gracilis} survival rates were
sensitive to temperature, resulting in an increase in plant cover under
increased temperature (Fig. 3e). In isolation, recruitment and survival
were insensitive to climate factors for \emph{H. comata} (Fig. 3f).
Survival and recruitment of \emph{P. smithii} were both sensitive,
negatively, to temperature and precipitation (Fig. 3g). \emph{P.
secunda} equilibrium cover was sensitive to the climate effects on
survival and recruitment, showing a negative effect on both vital rates
for increased precipition, but a strong positive effect on survival with
increased temperature (Fig. 3h). The climate impact of recruitment on
equilibrium cover was negative for precipitation and temperature
increases (Fig. 3h). At least two of three vital rates were sensitive to
climate for each species (Fig. 3).

Forecasts based on 1\% climate changes were extremely uncertain when we
considered model error and parameter uncertainty (Fig. 4). As expected
based on model validation (Table 1), QBM projections were more uncertain
than IPM projections for all species except \emph{P. smithiii} (Fig. 4).

@@

MOVE FIRST TWO PARAGRAPHS TO INTRO. MOVE FIG 5 and 6 TO METHODS AND
RESULTS

As a potential remedy to the ``data dearth'' problem, Queenborough et
al. (2011) and Freckleton et al. (2011), building on work by Taylor and
Hastings (2004), advocate a ``density-structured'' modeling approach.
Such models do not require individual level demographic data and can
adequately describe population dynamics (Queenborough et al. 2011). The
results from density-structured models are not as precise as those from
traditional population models, but the loss in precision is traded off
with a gain in data. The study by Queenborough et al. (2011) included
data from 500 fields (4 hectares each) in 49 farms, all collected by two
people in 6 weeks. This is far more data from a far greater spatial
extent than possible if measuring individual plant demography (in a
world of limited time and money, at least). The appeal of
density-structured approaches is clear.

However, at their core, density-structured models rely on individual
level data aggregated to a population level metric (e.g., density
classes or percent cover). This creates a potential problem if such
models are to be used in a climate change context because inidividuals
respond to climate, not populations (Clark et al. 2012). Are models
based on population level metrics as sensitive to climate as models
based on individual level metrics? Do these two types of models produce
consistent forecasts? Do we need detailed demographic data to forecast
the impacts of climate change? These are the questions we sought to
answer here.

@@

Using individual and aggregated forms of the same dataset, we were able
to directly compare a traditional demographic modeling approach to a
population model based on aggregated data. Our quad-based model (QBM) is
based on percent cover data and so is in the spirit of
density-structured models. In terms of each model's forecasting ability,
the IPM outperformed the QBM (Table 1). This is unsurprising since we
expected to lose some precision at the aggregated level. However, the
underwhelming performance of the QBM could call into the question
forecasts that differ from the IPM.

>>>>

To date, most empirical population modelers have relied on
individual-level data, with few attempts to exploit population-level
data. An important exception was an effort by Taylor and Hastings (2004)
to model the population growth rate of an invasive species to identify
the best strategies for invasion control. They used a
``density-structured'' model where the state variable is a discrete
density state rather than a continuous density measure. Such models do
not require individual level demographic data and can adequately
describe population dynamics.

Building on Taylor and Hastings (2004), Freckleton et al. (2011) showed
that density-structured models compare well to continuous models in
theory, and Queenborough et al. (2011) demonstrated the application of
such methods in a study on arable weeds. In particular, Queenborough et
al. (2011) provide empirical evidence that density-structured models are
capable of reproducing population dynamics, even if some precision is
lost when compared to fully continuous models. The study by Queenborough
et al. (2011) included data from 500 fields (4 hectares each) in 49
farms, all collected by two people in 6 weeks. This is far more data
from a far greater spatial extent than possible if measuring individual
plant demography (in a world of limited time and money, at least). The
appeal of density-structured approaches is clear. However, none of these
models included environmental covariates.

@@

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual-level data we fit models for survival, growth, and
recruitment for each species. At the quadrat-level we fit a single
regression model for population growth. We describe the statistical
models separately since fitting the models required different
approaches. All models contain five climate covariates that we chose
\emph{a priori}: ``water year'' precipitation at \emph{t}-1 (lagppt);
April through June precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and April through June temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year. We also include interactions among same-year
climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$
size interactions. Climate $\times$ size interactions are for climate
main effects only, that is we do not include interactions among size and
interacting climate effects.

We fit all models using a hierarchical Bayesian approach. The models are
fully descibed in Appendix A, so here we focus on the main process and
the model likelihood. For the likelihood models, $\textbf{y}^X$ is
always the relevant vector of observations for vital rate \emph{X}
($X = S, G, R, or P$ for survival, growth, recruitment, or population
growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating
whether a genet survives from \emph{t} to \emph{t+1}, or not.

@@

which means we are calculating growth (\emph{G}) for individuals that
survive (\emph{S}) from time \emph{t} to \emph{t+1} and adding in newly
recruited (\emph{R}) individuals of an average sized one-year-old genet
for the focal species. Our stastical model for recruitment (\emph{R},
described above) returns the number of new recruit produced per quadrat.
Following previous work (Adler et al. 2012, Chu and Adler 2015), we
assume that fecundity increases linearly with size
($R_j(v_{j},u_{j},\bar{\bold{w_{j}}}) = e^{u_j}R_j(v_{j},\bar{\bold{w_{j}}})$)
to incorporate the recruitment function in the spatially-implicit IPM.
\todo{Add spatial structure mention.}

@@

Our quad-based model (QBM) perfectly mirrors its statistical description
(Eqs. 8-9). We use the same approach for drawing parameter values as
described for the IPM. After drawing the appropriate parameter set, we
calculate the mean response (population cover at \emph{t+1} = $x_{t+1}$)
according to Eq. 8. We then make a random draw from a {[}0,1{]}
truncated lognormal distribution with mean equal to $x_{t+1}$ from Eq. 8
and the variance estimate from the fitted model. We can then iterate the
model forward by drawing a new parameter set (unique to climate year and
MCMC iteration) at each timestep.

@@

Our \emph{a priori} expectation is that the IPM will produce more
accurate and precise forecasts. This could be due to differences in
sample sizes leading to larger parameter uncertainty for the QBM, or due
to the QBM climate effects being weakly associated with one or more
vital rate climate effects at the individual level. To assess the impact
of sample size on QBM parameter uncertainty we refit the QBM statistical
model (Eqs. 8-9) after removing sets of 2, 5, 10, and 15 quadrats. We
fit 10 models at each level of quadrat removal (2, 5, 10, 15 quadrats),
removing a different randomly selected set of quadrats for each fit. We
calculated the standard deviation of climate main effects (pptlag, ppt1,
ppt2, TmeanSpr1, and TmeanSpr2) for each model and averaged those over
replicates within each set of quadrat removals. This allowed us to
regress sample size against parameter uncertainty.

To see if the QBM climate effects are correlated, or not, with climate
effects for each vital rate model for the IPM, we simply regressed the
QBM climate coefficients against each vital rate model's climate
coefficients and calculate Pearson's $\rho$. Strong correlations
indicate the QBM is capable of detecting climate effects associated with
individual vital rates. A weak correlation indicates the QBM ``misses''
the climate effect on a particular vital rate.

@@

The response of a population to climate change is a result of the
aggregate effects of climate on individual vital rates. Since the IPM
approach relies on vital rate regressions, we were able to quantify the
sensitivity of each vital rate in isolation and in pairwise
combinations. Across all species, climate covariates can have opposing
effects on different vital rates (Fig. 3e-h).

Growth was the most sensitive vital rate for all species, showing a
negative response to increased precipitation, and stronger positive
response to increased temperature, and a mostly positive response when
both climate factors are increased (Fig. 3e-h). \emph{B. gracilis}
survival rates were sensitive to temperature, resulting in an increase
in plant cover under increased temperature (Fig. 3e). In isolation,
recruitment and survival were insensitive to climate factors for
\emph{H. comata} (Fig. 3f). Survival and recruitment of \emph{P.
smithii} were both sensitive, negatively, to temperature and
precipitation (Fig. 3g). \emph{P. secunda} equilibrium cover was
sensitive to the climate effects on survival and recruitment, showing a
negative effect on both vital rates for increased precipition, but a
strong positive effect on survival with increased temperature (Fig. 3h).
The climate impact of recruitment on equilibrium cover was negative for
precipitation and temperature increases (Fig. 3h). At least two of three
vital rates were sensitive to climate for each species (Fig. 3).

Forecasts based on 1\% climate changes were extremely uncertain when we
considered model error and parameter uncertainty (Fig. 4). As expected
based on model validation (Table 1), QBM projections were more uncertain
than IPM projections for all species except \emph{P. smithiii} (Fig. 4).
IPM forecasts for \emph{P. smithiii} were very uncertain due to
overcompensating density dependence and nonlinearities recruitment that
get amplified by parameter uncertainty (Appendx C). Neither model was
capable of making forecasts of proportional cover change distinguishable
from 0 (i.e., no change in population state) when we included model
error and parameter uncertainty.

\subsubsection{Sources of uncertainty in the
QBM}\label{sources-of-uncertainty-in-the-qbm}

Sample size had a relatively weak effect on QBM climate parameter
uncertainty after the number of quadrats exceeded about 10 (Fig. 5).
Inverse-gaussian fits show that increasing sample size beyond the number
of quadrats we used will results in deminishing returns in terms of
parameter certainty (Fig. 5).

Climate effects estimated from the QBM are most correlated with climate
effects from the growth regression at the individual level (Fig. 6). In
no case does the QBM statistical model have strong correlations across
all three vital rates (Fig. 6). QBM climate effects were most weakly
correlated with those from individual-level recruitment models for
\emph{B. gracilis}, \emph{H. comata}, and \emph{P. secunda} (Fig.
6a,b,d). For \emph{P. smithii}, QBM climate effects showed no
correlation with the survival model effects (Fig. 6c).

@@

Population models built using individual-level data allow inference on
demographic processes, but their ability to forecast future population
states is limited by the spatial extent of the data. Population-level
data, like percent cover or discrete density states, are much easier to
collect of broad spatial extents, so populations built using such data
offer an appealing alternative to traditional population models
(Queenborough et al. 2011). However, at their core, density-structured
models rely on individual level data aggregated to a population level
metric. This creates a potential problem if such models are to be used
in a climate change context because inidividuals respond to climate, not
populations (Clark et al. 2012). Are models based on population level
metrics as sensitive to climate as models based on individual level
metrics? Do these two types of models produce consistent forecasts? Do
we need detailed demographic data to forecast the impacts of climate
change? These are the questions we sought to answer here.

@@

Using individual-level and population-level forms of the same dataset,
we were able to directly compare a traditional demographic modeling
approach to a population model based on percent cover data. Our
quad-based model (QBM) is based on percent cover data and so is in the
spirit of density-structured models. In terms of each model's
forecasting ability, the IPM outperformed the QBM (Table 1). This is
unsurprising since we expected to lose some precision at the aggregated
level. However, the underwhelming performance of the QBM could call into
the question forecasts that differ from the IPM.
